---
title: "3.2 Vector Indexing - Tutorial"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = TRUE)

library(ottr)
library(testthat)
```

---

At the end of this tutorial you should be able to:

* Index one or more elements of a vector using positional indexing
* Index one or more elements of a vector using named indexing
* Modify specific elements of a vector
* Write logical expressions with vectors
* Index a vector using logical indexing

---

## 3.2.1: Positional Indexing

---

We will use the same data as the last tutorial. As a reminder, a scientist has measuring the growth of wild-type (no mutations) E. coli in three conditions:
1. in the presence of the antibiotic kanamycin
2. in the presence of the antibiotic ampicillin
3. in the presence of no antibiotics

The OD measurements for 6 replicates of E. coli after 24 hours in the presence of no antibiotics are saved in the vector `no_antibiotic_OD`.

```{r}
no_antibiotic_OD <- c(1.32, 1.02, 1.48, 1.71, 1.23, 1.45)
```

The OD measurements for 3 replicates of E. coli after 24 hours in the presence of kanamycin are saved in the vector `kanamycin_OD`.

```{r}
kanamycin_OD <- c(0.05, 0.01, 0.03)
```

The OD measurements for 3 replicates of E. coli after 24 hours in the presence of ampicillin are saved in the vector `ampicillin_OD`.

```{r}
ampicillin_OD <- c(0.02, 0.01, 0.04)
```

Use positional indexing to retrieve the fifth element in the `no_antibiotic_OD` vector. Assign the value to a variable named "no_antibiotic_OD_5".
```{r}
# Write and execute your code here
# YOUR CODE HERE
```

```{r}
. = ottr::check("tests/Tutorial.W3.2.1.a.R")
```
**Check your answer by running the above block** 

Use positional indexing with a sequence to retrieve the second to the fifth elements of the `no_antibiotic_OD` vector. Assign the new vector to a variable named "no_antibiotic_OD_2_to_5".
```{r}
# Write and execute your code here
# YOUR CODE HERE
```

```{r}
. = ottr::check("tests/Tutorial.W3.2.1.b.R")
```
**Check your answer by running the above block** 

Retrieve elements 2, 3, and 6 from the `no_antibiotic_OD` vector. Assign the new vector to a variable named "no_antibiotic_OD_2_3_6".
```{r}
# Write and execute your code here
# YOUR CODE HERE
```

```{r}
. = ottr::check("tests/Tutorial.W3.2.1.c.R")
```
**Check your answer by running the above block** 

Output all three elements in the `ampicillin_OD` vector in reverse order. Assign the new vector to a variable named "amp_OD_rev".
```{r}
# Write and execute your code here
# YOUR CODE HERE
```

```{r}
. = ottr::check("tests/Tutorial.W3.2.1.d.R")
```
**Check your answer by running the above block** 

Output all of the elements in `no_antibiotic_OD` vector EXCEPT for the second and fourth elements (use negative indices). Assign the new vector to a variable named "no_antibiotic_OD_1_3_5_6".
```{r}
# Write and execute your code here
# YOUR CODE HERE
```

```{r}
. = ottr::check("tests/Tutorial.W3.2.1.e.R")
```
**Check your answer by running the above block** 

---

## 3.2.2: Named Indexing & Editing Vectors

---

The scientist has 8 mutant strains of E. coli that grow in the presence of no antibiotic that they hypothesize are resistant to antibiotics. They want to check if any of them are resistant to kanamycin or ampicillin, in other words, if they grow in the presence of kanamycin or ampicillin.

The scientist grows the 8 mutant strains in *kanamycin* and checks the OD after 24 hours. 

|    | mut_A | mut_B | mut_C | mut_D | mut_E | mut_F | mut_G | mut_H | 
| OD | 1.24  | 0.02  | 0.13  | 1.45  | 1.36  | 0.04  | 0.01  | 1.19  |

The scientist grows the 8 mutant strains in *ampicillin* and checks the OD after 24 hours. 

|    | mut_A | mut_B | mut_C | mut_D | mut_E | mut_F | mut_G | mut_H | 
| OD | 0.06  | 0.03  | 1.57  | 0.04  | 1.48  | 0.02  | 1.77  | 0.04  |

```{r}
mut_kan_OD <- c("mut_A"=1.24, "mut_B"=0.02, "mut_C"=0.13, "mut_D"=1.45, "mut_E"=1.36, "mut_F"=0.04, "mut_G"=0.01, "mut_H"=1.19)

mut_amp_OD <- c("mut_A"=0.06, "mut_B"=0.03, "mut_C"=1.57, "mut_D"=0.84, "mut_E"=1.48, "mut_F"=0.02, "mut_G"=1.77, "mut_H"=0.04)
```

Retrieve the OD measurement after 24 hours on ampicillin for mutant G using named indexing. Assign the value to a variable named "mut_G_amp_OD".
```{r}
# Write and execute your code here
# YOUR CODE HERE
```

```{r}
. = ottr::check("tests/Tutorial.W3.2.2.a.R")
```
**Check your answer by running the above block** 

Retrieve the OD measurements after 24 hours on ampicillin for mutant D, mutant F, and mutant H using named indexing. Assign the new vector to a variable named "mut_D_F_H_amp_OD".
```{r}
# Write and execute your code here
# YOUR CODE HERE
```

```{r}
. = ottr::check("tests/Tutorial.W3.2.2.b.R")
```
**Check your answer by running the above block** 

Retrieve the OD measurements after 24 hours on ampicillin for mutants H and A (in that order) using named indexing. Assign the new vector to a variable named "mut_H_A_amp_OD".
```{r}
# Write and execute your code here
# YOUR CODE HERE
```

```{r}
. = ottr::check("tests/Tutorial.W3.2.2.c.R")
```
**Check your answer by running the above block** 

When copying the OD measurements the scientist made a mistake. The OD measurement for mutant C grown on kanamycin is actually 0.03 not 0.13. Replace the value at "mut_C" in `mut_kan_OD` with 0.03 (use named indexing).
```{r}
# Write and execute your code here
# YOUR CODE HERE
```

```{r}
. = ottr::check("tests/Tutorial.W3.2.2.d.R")
```
**Check your answer by running the above block** 

Later on, the scientist found that the 2nd wild-type replicate grown on no antibiotic was contaminated. Remove the value at index 2 from the `no_antibiotic_OD` vector.
```{r}
# Write and execute your code here
# YOUR CODE HERE
```

```{r}
. = ottr::check("tests/Tutorial.W3.2.2.e.R")
```
**Check your answer by running the above block** 

---

## 3.2.3: Logical Indexing

---

The scientist thinks that an OD of 1 should indicate that E. coli can grow in a given condition. If this is true, all of the wild-type replicates grown without antibiotic should have an OD > 1. 

Write a logical expression vector to indicates for each element in `no_antibiotic_OD` whether it is greater than 1. Assign the new logical vector to a variable named "no_antibiotic_OD_above_1".
```{r}
# Write and execute your code here
# YOUR CODE HERE
```

```{r}
. = ottr::check("tests/Tutorial.W3.2.3.a.R")
```
**Check your answer by running the above block** 

Output a vector that indicates for each element in `mut_amp_OD` whether it is less than 0.02 or greater than 1. Assign the new logical vector to a variable named "mut_amp_OD_less_0.02_or_above_1".
```{r}
# Write and execute your code here
# YOUR CODE HERE
```

```{r}
. = ottr::check("tests/Tutorial.W3.2.3.b.R")
```
**Check your answer by running the above block** 

Use a logical expression inside of square brackets for the rest of the questions in this tutorial.

The scientist decides that 1 is a good threshold for growth.

Can wild-type E. coli grown on kanamycin? Use logical indexing to retrieve the elements in `kanamycin_OD` that are greater than 1. Assign the result of the logical indexing to a variable named "wild_type_growth_on_kan".
```{r}
# Write and execute your code here
# YOUR CODE HERE
```

```{r}
. = ottr::check("tests/Tutorial.W3.2.3.c.R")
```
**Check your answer by running the above block** 

Can wild-type E. coli grown on ampicillin? Use logical indexing to retrieve the elements in `ampicillin_OD` that are greater than 1. Assign the result of the logical indexing to a variable named "wild_type_growth_on_amp".
```{r}
# Write and execute your code here
# YOUR CODE HERE
```

```{r}
. = ottr::check("tests/Tutorial.W3.2.3.d.R")
```
**Check your answer by running the above block** 

Which of the 8 mutants are kanamycin resistant? Use logical indexing to retrieve the elements in `mut_kan_OD` that have an OD > 1. Assign the result of the logical indexing to a variable named "kan_resistant_mutants".
```{r}
# Write and execute your code here
# YOUR CODE HERE
```

```{r}
. = ottr::check("tests/Tutorial.W3.2.3.e.R")
```
**Check your answer by running the above block** 

Which of the 8 mutants are ampicillin resistant? Use logical indexing to retrieve the elements in `mut_amp_OD` that have an OD > 1. Assign the result of the logical indexing to a variable named "amp_resistant_mutants".
```{r}
# Write and execute your code here
# YOUR CODE HERE
```

```{r}
. = ottr::check("tests/Tutorial.W3.2.3.f.R")
```
**Check your answer by running the above block** 

Looking at the OD values of wild-type E. coli grown in ampicillin and kanamycin, the scientist decides that an OD < 0.1 means that an E. coli strain can't grown in certain condition.

Do any of the 8 mutants have an OD measurement on ampicillin that shows an unclear result? Use logical indexing to retrieve the elements in `mut_amp_OD` that have an OD > 0.1 and < 1. Assign the result of the logical indexing to a variable named "amp_inconclusive_mutants".
```{r}
# Write and execute your code here
# YOUR CODE HERE
```

```{r}
. = ottr::check("tests/Tutorial.W3.2.3.g.R")
```
**Check your answer by running the above block** 

---